Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-02 text/discussion.qmd
working through discussion

The discussions of correlation and homophily by @RN4975 describe the logic of The Market Segment model but also the underlying institutional logic of the enrollment management industry. The logic of correlation argues that colleges should find students who look like students who enrolled at your institution in the past. The logic of homophily argues that there is a cultural match between institutions and students, for example, that  Catholic universities compete with Catholic universities for Catholic students [@RN4982, p. 84]. Therefore, @RN4982 recommends that colleges recruit from the same schools and communities that are popular with peer colleges. The operating logics of correlation and homophily are highly consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. The also find great overlap in the schools visited by Villanova and Notre Dame, with a disproportionate focus on visits to Catholic schools. When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

The Market Segment Model begins with the status attainment model, that student demand for college is a function of parental education and parental income. @RN4982 demonstrated this argument by showing the correlation between student SAT score-sending behavior and measures of class. The thesis of the Market Segment Model is that that households of particular class and college aspirations are likely to live in particular parts of a metropolitan area and that these geographic territories can be meaningfully captured by Geomarkets. Although they appear coarse by contemporary standards, Geomarkets were a landmark innovation in Geographic segmentation. The Market Segment Model recommends that colleges should identify their core student market segment -- local, in-state, regional, or national -- identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile (Appendix A), which is a standard output of EPS software. When considering new Geomarkets to target, colleges should utilize the logic of homophily by identifying Geomarkets that are popular with peer colleges. This information is contained in the Institutional Profile (Appendix A). Technologies that target target customers based on geography leverage racial and class segregation [@RN4775; @RN4849; @RN4975]. The Market Segment Model and Geomarkets were created based on 1980 SAT score-sending patterns, contributing to a "ratchet effect" [@RN4851]. The structural barriers (e.g., segregation, slavery, Jim Crow) that caused inequality in 1980 student demand are amplified because the Market Segment Model recommends that colleges focus recruiting resources on localities that already have high student demand.

@@

These analyses contribute to scholarship about how students are sorted into colleges [IMPROVE TRANSITION]. The status attainment model argues that college destination is a function of parental education and occupation [@RN1273; @RN1271; @RN1272]. Indeed, the Market Segment Model can be seen as a market research application of the status attainment model. @RN5033 finds that Asian American students -- particularly children of Chinese, Indian, Korean, and Vietnamese immigrants -- tend to have high levels of educational achievement even if their parents do not, while the educational achievement of later-generation Asian Americans conforms to status attainment theory. In our analyses of prospects included in student list purchases, Asian American prospects from poor Geomarkets tended to be first-generation students while Asian American prospects from affluent Geomarkets tended to have parents with a BA. Drawing from @RN5033, one explanation for this finding is that Asian students living in low-income Geomarkets tend to have parents who immigrated to the U.S. whereas Asian students living in affluent Geomarkets are more likely to have parents who were born in the U.S.

The cultural capital model explains a process by which upper and upper-middle class families sort themselves into selective colleges by providing their children with the pedigree (academic, extracurricular) valued by selective colleges, information about how to navigate the admissions gauntlet, and social networks that provide an inside track [@RN1228; @RN1241; @RN4991; @RN1814]. The cultural capital model explains how affluent households maintain a disproportionate enrollment share at selective institutions in an era of holistic admissions that has the purported mission of increasing racial and class diversity. <!--Amidst concern about discrimination against Asian students at Ivy Plus institutions, @RN4991 shows how private admissions consultants coach Asian American students to......NEED TO REVISE THIS .-->

Both the status attainment model and the cultural capital model are demand-side explanations for why education is a "social sieve" [@RN2377; @RN5046] that allows for a modicum of social mobility while maintaining a much larger flow of intergenerational class transmission [@RN861]. The relative sizes of these flows can be observed in @RN4722, which obtained parental income from federal income tax returns for every college in the U.S. The disproportionate enrollment share of high income families at selective private colleges -- and even at most public research universities -- is staggering. These patterns are consistent with the observation by @RN1297[p. 241-242] that educational credentials are property that "support their holders'\ldots claims to monopolize socially and economically advantageous positions."

Complementing the cultural capital explanation of how students are sorted into colleges are the supply-side explanations of the credentialism literature and the more recent literature on enrollment management. The credentialism literature recognizes that colleges have a financial incentive for educational credentials to be required for socially and economically advantageous positions [@RN865; @RN861; @RN880; @RN869]. The enrollment management literature describes which prospects colleges want to enroll and what colleges do to get these students, at different stages of the enrollment funnel [@RN3519; @RN4324; @RN3522; @RN4407; @RN4758; @RN3544; @RN3547; @cottom2017lower]. @RN3519 and @RN4321 describe a tacit arrangement between high school guidance counselors on the demand side and college admissions officers on the supply side. High school guidance counselors at well-resourced schools, especially private schools, are motivated to give their students a competitive advantage in admissions. Meanwhile, college admissions officers are motivated to enroll students who can afford full tuition price and are likely to donate in the future. These mutually beneficial desires are consummated by recognzing that upper and upper-middle class applicants satisfy the extracurricular needs of the college. The band needs oboists. The lacrosse team needs players. Quantitative analyses of visits from colleges to high schools are consistent with @RN3519; selective private colleges and public research universities devote most of their recruiting resources on courting students from privileged schools and communities [@RN5023; @RN4758]. However, the extant enrollment management literature is flawed in that it views recruiting as something that is done by individual colleges.

We argue that existing explanations how students are sorted into colleges miss an important supply-side mechanism, third-party vendors that sort students on behalf of colleges. Based on a snapshot of student demand from 1980, @RN4982 created the Market Segment Model and Geomarkets. @RN4982 argues that demand for higher education is correlated with class, ingoring the historical structural barriers that produced class- and race-based inequality in student demand. In itself, this explanation is basically the same as the status attainment model. However, the College Board goes further by commodifying the Market Segment Model into products that tell colleges which Geomarkets to target. Therefore, third-party products encourage supply-side enrollment management behaviors to amplify historical structural inequalities observed on the demand-side. Geomarkets were incorporated into the Student Search Service product. We show that that excluding low-income, non-white Geomarkets from student list purchases results in the disproportionate exclusion of first-generation, non-white students with strong test scores.

@@

The growing salience of third-party vendors warrants a reconfiguration of the organizational field that is salient to college access. @RN527[p. 148] define organizational fields as, "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services or products" and that "the virtue of this unit of analysis is that it directs our attention\ldots to the totality of relevant actors." Sociological literatures salient to college access devote substantive attention to schools, colleges, families, and communities. @RN3547 shows how local external stakeholders intervene on admissions standards. @RN4991 describes the role of independent admissions consultants hired by families as third-party vendors that engage on the demand-side. College access scholarship from economics analyzes the effects of policies (district/state/federal) and the effects of information interventions [CITE HOXBY TYPE]. None of these explanations seriously consider the growing role of third-party vendors, which are increasingly owned by private equity interests and offer software-as-service platforms that perform core functions for schools and colleges.

@list_biz describes four key dynamics in the enrollment management (EM) industry, with a focus on the market for student list data. First, EM consulting firms are central to the creation and implementation of recruiting campaigns [@RN5027]. When we issued records requests to public universities about their student list purchases, at least 50\% of universities indicated that they outsourced student list purchases to an EM consultancy []. In many cases, the university was buying names each year but no university employee had knowledge about which student list vendors they were buying names from. This reliance on EM consultancies is partially explained by the high level of burnout and employee turnover in the EM profession [@hoover_em].

Second, technological advances created means of identifying and serving prospects. College Board began selling the contact information of test-takers in 1972 [@belkin2019-studata]. ACT followed suit and the two testing companies enjoyed a near duopoly for several decades. This business can be described as list-based lead generation based on the direct-mail model. "Free" college search engines (e.g., Cappex, Niche) yielded news sources of student list data. Another source of student list data is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than used by more than 10 million K-12 students and by 40% ofUS high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform, which allows colleges to target Naviance users while they are on the platform. *Intersect* is an example of behavioral-based -- as opposed to list-based -- targeting.

The third dynamic in the EM industry has been the growth of test-optional movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company, failing miserably, and being acquired by Nexus Capital. [ADD A COUPLE SENTENCES ABOUT COLLEGE BOARD; MAYBE GEODEMOGRAPHIC SEGMENT]

The fourth dynamic is the transformation from owner-operated to private-equity owned firms and the blurring between consultancy and software vendors. The 1980s and 1990s was a period of market entry in EM, mostly consistent of college admissions professionals or professors who decided to hang up a shingle [@RN5027]. The 2010s were a period of acquisitions and concentration. After regulatory scrutiny pushed private-equity interests out of for-profit college market [@RN5022], private equity investors saw value in acquiring firms that provided services to direct providers. Drawing from resource dependence theory, @list_biz describe the proliferation of horizontal acquisitions grow market share and vertical acquisitions designed to increase customer reliance.

EAB exemplifies the the trend towards private equity ownership and industry concentration. EAB is the preeminent enrollment management (EM) consulting firm, serving more than 2,100 colleges and offering software-as-service products along the domains of recruitment, pricing and financial aid, student success, and advancement [@about_eab]. The origins of EAB are traced to Bill Royall, who founded Royall & Company in 1983 to provide direct marketing and fundraising for Republican political campaigns [@jump_2020]. Royall did not sign its first colege client for several years, but EM became the primary focus by 1995. Royall was acquired by the health tech firm Advisory Board Company for \$850 million in 2015 and then by Vista Equity Partners for \$1.5 billion in 2017, renaming it EAB [@drew_hansen_2017].

>>>>

The discussion of homophily by @RN4975 describe the logic of The Market Segment model but also an underlying institutional logic of the enrollment management industry. The logic of homophily argues that there is a natural cultural match between students, that colleges should find students who look like students who enrolled at their institution in the past, and that colleges should recruit from the same schools and communities that are popular with peer colleges. We suggest that homophily is the operating logic most consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. The also find great overlap in the schools visited by Villanova and Notre Dame, with a disproportionate focus on visits to Catholic schools. When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

The Market Segment Model begins with the status attainment model, that student demand for college is a function of parental education and parental income. @RN4982[chapter 3] demonstrated this argument by showing the correlation between student SAT score-sending behavior and measures of class. The thesis of the Market Segment Model is that that households of particular class and college aspirations are likely to live in particular parts of a metropolitan area and that these geographic territories can be meaningfully captured by Geomarkets. Although Geomarkets appear coarse by contemporary standards, they were a landmark innovation in Geographic segmentation. The Market Segment Model recommends that colleges should identify their core student market segment -- local, in-state, regional, or national -- identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile (Appendix A), a standard ouput in EPS software. When considering new Geomarkets to target, colleges should utilize the logic of homophily by identifying Geomarkets that are popular with peer colleges. This information is contained in the Institutional Profile (Appendix A), also included in EPS software.

@@

This article adds a new perspective to existing explanations about how students are sorted into colleges. The status attainment model argues that college destination is a function of parental education and occupation [@RN1273; @RN1271; @RN1272]. The Market Segment Model is the status attainment model applied to geodemographic market research [CITE]. @RN5033 finds that the educational achievement of Asian American students whose parents are immigrants tend to have high educational achievement even if their parents do not, while the educational achievement of later-generation Asian Americans conforms to status attainment theory. In our analyses of prospects included in student list purchases, Asian American prospects from poor Geomarkets tended to be first-generation students while Asian American prospects from affluent Geomarkets tended to have parents with a BA. Drawing from @RN5033, one explanation for this finding is that Asian students living in low-income Geomarkets tend to have parents who immigrated to the U.S. whereas Asian students living in affluent Geomarkets are more likely to have parents who were born in the U.S.

The cultural capital model explains a process by which upper and upper-middle class families sort themselves into selective colleges by providing their children with the pedigree (academic, extracurricular) valued by selective colleges, information about how to navigate the admissions gauntlet, and social networks that provide an inside track [@RN1228; @RN1241; @RN4991; @RN1814]. This model explains how affluent households maintain a disproportionate enrollment share at selective institutions in an era of holistic admissions ostensibly designed to increase racial and class diversity [@RN4991]. Both the status attainment model and the cultural capital model are demand-side explanations for why education is a "social sieve" [@RN2377; @RN5046] that allows for a modicum of mobility while maintaining a much larger flow of intergenerational class transmission [@RN861]. @RN4722 -- who obtained parental income from federal income tax returns for every U.S. college -- show that the disproportionate enrollment share of high income families at selective private colleges -- and even at most public research universities -- is staggering. These patterns are consistent with the observation by @RN1297[p. 241-242] that educational credentials are property that "support their holders'\ldots claims to monopolize socially and economically advantageous positions."

Supply-side explanations of credentialism and more recent scholarship on enrollment management complement the cultural capital model of sorting students to colleges. The credentialism literature recognizes that colleges have a financial incentive for educational credentials to determine the competition for socially and economically advantageous positions [@RN865; @RN861; @RN880; @RN869] and this credentials arms race favors affluent households. Scholarship on enrollment management describes which students colleges want to enroll and what colleges do to attract these students at different stages of the enrollment funnel [@RN3519; @RN4324; @RN3522; @RN4407; @RN4758; @RN3544; @RN3547; @cottom2017lower]. @RN3519 and @RN4321 describe a tacit arrangement between high school guidance counselors on the demand side and college admissions officers on the supply side. Counselors at well-resourced schools, especially private schools, are motivated to give their students a competitive advantage in admissions. Meanwhile, college admissions officers are motivated to enroll students who can afford full tuition price and are likely to donate in the future. These mutually beneficial desires are consummated by recognzing that upper and upper-middle class applicants satisfy the extracurricular needs of the college. The orchestra needs oboists. The lacrosse team needs players. Quantitative analyses of visits from colleges to high schools are consistent with @RN3519; selective private colleges and public research universities devote most of their recruiting resources on courting students from privileged schools and communities [@RN5023; @RN4758]. However, these analyses conceive of recruiting as something that is done by individual colleges.

Existing explanations of how students are sorted into colleges miss an important supply-side mechanism, third-party vendors that sort students on behalf of colleges. @RN4982 created the Market Segment Model and Geomarkets based on a snapshot of student demand from 1980. @RN4982 argues that demand for higher education is correlated with class, ignoring the historical structural barriers that produced class- and race-based inequality in 1980 student demand. In itself, the Market Segment Model is an unremarkable social science depiction. However, the College Board incscribed Geomarkets and the Market Segment Model into EPS software that told colleges which Geomarkets and high schools to target. Technologies that target target customers based on geography inevitably leverage racial and class segregation [@RN4775; @RN4849; @RN4975]. Furthermore, technologies that use the logic of predictive analytics -- identify correlates in past cases to make predictions for future cases -- are prone to "pernicious feedback loops" [@RN4849]. The structural barriers (e.g., segregation, slavery, Jim Crow) that caused historic place-based inequality in student demand are amplified because EPS recommends that colleges focus recruiting resources on localities that already have high student demand for peer colleges. Geomarkets were subsequently incorporated into the Student Search Service product. We show that that excluding low-income, non-white Geomarkets from student list purchases results in the disproportionate exclusion of first-generation, non-white students with strong test scores.

@@

The growing salience of third-party vendors warrants a reconfiguration of the organizational field salient to college access. @RN527[p. 148] define organizational fields as, "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services or products" and that "the virtue of this unit of analysis is that it directs our attention\ldots to the totality of relevant actors." Sociological literatures salient to college access devote substantive attention to schools, colleges, families, and communities.College access scholarship from economics often analyzes the effects of district/state/federal [e.g., @RN5043] policies and the effects of information interventions [@RN4018]. These lines of research do not consider the growing role of third-party vendors, which are increasingly owned by private equity interests and offer software-as-service platforms that perform core functions for schools and colleges.^[ @RN4991 describes the role of independent admissions consultants hired by families as third-party vendors that engage on the demand-side.]

@list_biz describes four key dynamics in the enrollment management (EM) industry, with a focus on the market for student list data. First, EM consulting firms are central to the creation and implementation of recruiting campaigns [@RN5027]. The top two EM consultancies are Ruffalo Noel Levitz, which claims to serve 1,900 colleges annually [@rnlabout_2025], and EAB, which claims to serve 2,100 colleges [@about_eab]. College reliance on EM consultancies is partially explained by the high level of burnout and employee turnover in the EM profession [@hoover_em]. When we issued records requests to public universities about their student list purchases, at least 50\% of universities indicated that they outsourced student list purchases to an EM consultancy. In many cases, the university was buying names each year but no university employee had knowledge about which vendors they were buying from or which search filters they were utilizing.

Second, technological advances creates new means of identifying and serving prospects. College Board began selling the contact information of test-takers in 1972 [@belkin2019-studata]. ACT followed suit and the two testing companies enjoyed a near duopoly for several decades. This business can be described as list-based lead generation based on the direct-mail model. "Free" college search engines (e.g., Cappex, Niche) yielded news sources of student list data. Another data source is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than used by more than 10 million K-12 students and by 40% ofUS high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform, which allows colleges to target Naviance users while they are on the platform. *Intersect* is an example of behavioral-based targeting.

The third dynamic in the EM industry has been the growth of test-optional movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company [CITE], failing [], and being acquired by Nexus Capital in 2024 [CITE]. Over the last two decades, the testing companies developed new search filters that explicitly incorporated predictive analytics -- for example College Board's Geodemographic Segment filters [CITE] and ACT's predicted probability of enrollment filter [CITE] -- developed free college search engines, and entered the enrollment management consulting market more aggessively. College Board has weathered the test-optional movement thanks to robust revenues from the Advanced Placement product, but the decline in PSAT/SAT/PreACT/ACT test-takers has undermined their oligopoly in the market for student list data.

The fourth dynamic is the transformation from owner-operated to private-equity owned firms and erosion of the distinction between consultancy and software vendor. The 1980s and 1990s was a period of market entry in EM, when college admissions professionals or professors decided to hang up a shingle [@RN5027]. The 2010s were a period of acquisitions and concentration. After regulatory scrutiny pushed private-equity interests out of for-profit college market [@RN5022], private equity investors find value in acquiring firms that provided services to direct providers. Drawing from resource dependence theory, @list_biz describe the proliferation of horizontal acquisitions to grow market share and vertical acquisitions designed to make customers more dependent on particular recruiting products.

EAB exemplifies the the trend towards private equity ownership and industry concentration. EAB is the preeminent EM consultancy, offering software-as-service products along the domains of recruitment, pricing and financial aid, student success, and advancement [@about_eab]. The origins of EAB trace to Bill Royall, who founded Royall & Company in 1983 to provide list-based marketing and fundraising for Republican political campaigns [@jump_2020]. Royall did not sign its first colege client for several years, but EM became the primary focus by 1995. Royall was acquired by the health tech firm Advisory Board Company for \$850 million in 2015 and then by Vista Equity Partners for \$1.5 billion in 2017, renaming it EAB [@drew_hansen_2017].
